\thetitle

### 2020-11-03 sun_phases/spin_qudit_tomo.tex
better explain the computation of the error scale

A choice of measurement axes $V=\set{\v v}$ determines a set of projectors $\v\Pi_V \equiv \set{\Pi_{\v v\mu}}$ that will be measured in an experiment.
By flattening qudit operators $X = \sum_{\mu,\nu} X_{\mu\nu} \op{\mu}{\nu}$ into the $d^2$-component vectors $\oket{X} \equiv \sum_{\mu,\nu} X_{\mu\nu} \ket{\mu\nu}$, we can construct the {\it measurement matrix}
\begin{align}
  M_V \equiv \sum_{\v v,\ell} \oket{\Pi_{\v v\mu}} \bra{\v v\mu},
  \label{eq:meas_mat}
\end{align}
whose columns are flattenings of the operators in $\v\Pi_V$.
A necessary and sufficient condition for $V$ to allow for full state tomography is that the projectors in $\v\Pi_V$, or equivalently the columns of $M_V$, to span the entire ($d^2$-dimensional) space of operators on a $d$-level spin qudit.
In this case $M_V$ must be full rank, with $d^2$ nonzero singular values.
Indexing these singular values $M^V_k$ and the corresponding (normalized) right singular vectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$, we can construct the orthonormal qudit operators
\begin{align}
  Q^V_k \equiv \sum_j q^V_{kj} \Pi_j,
  &&
  \v q^V_k \equiv \f{\v x^V_k}{M^V_k},
\end{align}
where for shorthand we use a combined index $j=\p{\v v_j,\mu_j}$ to specify both a measurement axis $\v v_j$ and a spin projection $\mu_j$, which identify the projector operator $\Pi_j\equiv \Pi_{\v v_j\mu_j}$.
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j$ of the expectation values $\bk{\Pi_j}_\rho$, an empirical estimate $\tilde\rho_V$ of $\rho$ is then
\begin{align}
  \tilde\rho_V \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j q^V_{kj} E_j
  \approx \sum_j q^V_{kj} \bk{\Pi_j}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The singular values of $M_V$ allow us to make concrete statements about the statistical error between the empirical estimate $\tilde\rho_V$ and the true state $\rho$.
Assume, for example, that the estimates $E_j$ are equal to $\bk{\Pi_j}_\rho$ up to uncorrelated noise with variance at most $\epsilon^2$, i.e.
\begin{align}
  E_j = \bk{\Pi_j}_\rho + \epsilon_j,
  &&
  \bbk{\epsilon_j \epsilon_{j'}} \le \epsilon^2 \delta_{jj'},
\end{align}
where $\set{\epsilon_j}$ are independent random variables and we use the double brackets $\bbk{\cdot}$ to denote statistical averaging.
The mean squared error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then
\begin{align}
  \Bbk{\abs{\tilde\rho^V_k-\rho^V_k}^2}
  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^*\p{\tilde\rho^V_k-\rho^V_k}} \\
  &= \sum_{j,j'} \p{q^V_{kj}}^* q^V_{kj'}\,
  \bbk{\epsilon_j \epsilon_{j'}} \\
  &\le \sum_j \abs{q^V_{kj}}^2 \epsilon^2
  = \p{\f{\epsilon}{M^V_k}}^2.
\end{align}
Using the fact that the operators $Q^V_k$ are orthonormal, we can therefore bound the mean squared 2-norm (Euclidean) distance between $\tilde\rho_V$ and $\rho$ as
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}_2^2}
  = \sum_k \Bbk{\abs{\tilde\rho^V_k - \rho^V_k}^2}
  \le \epsilon^2 \S_V^2,
  \label{eq:bound}
\end{align}
where the {\it error scale} $\S_V$ of $V$, defined by
\begin{align}
  \S_V \equiv \norm{M_V^{-1}}_2 = \sqrt{\sum_k\f1{\p{M^V_k}^2}},
  \label{eq:scale}
\end{align}
is the 2-norm of the pseudoinverse $M_V^{-1}$ of $M_V$.
Note that the bound in Eq.~\eqref{eq:bound} only holds if the measurement matrix $M_V$ is nonsingular, as otherwise some of the operators $Q^V_k$ that we used to expand $\rho$ and $\tilde\rho_V$ are undefined, reflecting the fact that if $M_V$ is singular then measuring spin projections along all axes in $V$ does not provide sufficient information to reconstruct arbitrary quantum states.

The complexity of computing the error scale $\S_V$ is determined by the need to
\begin{enumerate*}
\item construct the measurement matrix $M_V$, and
\item compute the singular values of $M_V$.
\end{enumerate*}
These tasks can be drastically simplified by use of the transition operators.
Specifically, the orthogonality of transition operators with different degrees, i.e.~$\obk{T_{\v v\ell,0}|T_{\v v',\ell',0}} = 0$ for $\ell\ne\ell'$, implies that $M_V$ has a block-diagonal structure in the transition operator basis, with rectangular blocks
\begin{align}
  M_{V\ell} \equiv U_\ell \sum_{\v v} \oket{T_{\v v\ell,0}} \bra{\v v},
  &&
  U_\ell \equiv \sum_{m=-\ell}^\ell \ket{m} \obra{T_{\ell m}},
\end{align}
that we can use to expand
\begin{align}
  \S_V^2 = \sum_\ell \S_{V\ell}^2,
  &&
  \S_{V\ell} \equiv \norm{M_{V\ell}^{-1}}_2,
\end{align}
where $M_{V\ell}^{-1}$ is the pseudoinverse of $M_{V\ell}$.
By the expansion $T_{\v v\ell,0} = R\p{\v v} T_{\ell,0} R\p{\v v}^\dag$, the $\p{2\ell+1}\times\abs{V}$ matrix $M_{V\ell}$ can be constructed in $O(\abs{V}d^3)$ time, and computing its singular values takes at most $O(\abs{V}d^2)$ time \footnote{The calculation of $\S_{V\ell}$ can be sped up by the fact that its value is independent of $d$ (as long as $d\ge\ell+1$), so we can essentially set $d=\ell+1$ for the purposes of computing $\S_{V\ell}$}.
Altogether, if we assume that $\abs{V}\sim d$ then computing $\S_V$ to certify the bound in Eq.~\eqref{eq:bound} takes $O(d^5)$ serial or $O(d^4)$ parallel runtime (see Figure [\red{REF}]).  [\red{todo: add figure showing runtimes as a function of $d$ with $\abs{V}=2d-1$; modify text appropriately after making the figure}].

[\red{todo: add a careful derivation of the error bound that assumes $n$ measurements of spin projection along each axis into an Appendix.  Briefly discuss the result of this derivation here.}]

@@

The reconstruction error bound in Eq.~\eqref{eq:bound} provides a straightforward answer: the number of measurement axes should be chosen to minimize the product $\epsilon^2\S_V^2$, which bounds the expected error in the estimate $\tilde\rho_V$ of an unknown state $\rho$.
If there are a total of $N$ spin projection measurements onto $r$ quantization axes, then $\sim N/r$ measurements are used to estimate spin projection onto any given axis.
The variance in estimates of spin projection is then $\epsilon^2\sim r/N$, or simply $\epsilon^2\sim r$ at a fixed total number of measurements $N$.
Altogether, the number of measurement axes should be chosen in such a way as to minimize $r\S_V^2$ with $\abs{V}=r$.

@@

>>>>

A choice of measurement axes $V=\set{\v v}$ determines a set of projectors $\v\Pi_V \equiv \set{\Pi_{\v v\mu}}$ that will be measured in an experiment.
By flattening qudit operators $X = \sum_{\mu,\nu} X_{\mu\nu} \op{\mu}{\nu}$ into the $d^2$-component vectors $\oket{X} \equiv \sum_{\mu,\nu} X_{\mu\nu} \ket{\mu\nu}$, we can construct the {\it measurement matrix}
\begin{align}
  M_V \equiv \sum_{\v v,\ell} \oket{\Pi_{\v v\mu}} \bra{\v v\mu},
  \label{eq:meas_mat}
\end{align}
whose columns are flattenings of the operators in $\v\Pi_V$.
A necessary and sufficient condition for $V$ to allow for full state tomography is that the projectors in $\v\Pi_V$, or equivalently the columns of $M_V$, to span the entire ($d^2$-dimensional) space of operators on a $d$-level spin qudit.
In this case $M_V$ must be full rank, with $d^2$ nonzero singular values.
Indexing these singular values $M^V_k$ and the corresponding (normalized) right singular vectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$, we can construct the orthonormal qudit operators
\begin{align}
  Q^V_k \equiv \sum_j q^V_{kj} \Pi_j,
  &&
  \v q^V_k \equiv \f{\v x^V_k}{M^V_k},
\end{align}
where for shorthand we use a combined index $j=\p{\v v_j,\mu_j}$ to specify both a measurement axis $\v v_j$ and a spin projection $\mu_j$, which identify the projector operator $\Pi_j\equiv \Pi_{\v v_j\mu_j}$.
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j$ of the expectation values $\bk{\Pi_j}_\rho$, an empirical estimate $\tilde\rho_V$ of $\rho$ is then
\begin{align}
  \tilde\rho_V \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j q^V_{kj} E_j
  \approx \sum_j q^V_{kj} \bk{\Pi_j}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The singular values of $M_V$ allow us to make concrete statements about the statistical error between the empirical estimate $\tilde\rho_V$ and the true state $\rho$.
Assume, for example, that the estimates $E_j$ are equal to $\bk{\Pi_j}_\rho$ up to uncorrelated noise with variance at most $\epsilon^2$:
\begin{align}
  E_j = \bk{\Pi_j}_\rho + \epsilon_j,
  &&
  \bbk{\epsilon_j \epsilon_{j'}} \le \epsilon^2 \delta_{jj'},
\end{align}
where $\set{\epsilon_j}$ are independent random variables and we use the double brackets $\bbk{\cdot}$ to denote statistical averaging (we discuss a more complicated and realistic scenario that leads to the same general conclusions in Appendix \ref{sec:error}).
The mean squared error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then
\begin{align}
  \Bbk{\abs{\tilde\rho^V_k-\rho^V_k}^2}
  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^*\p{\tilde\rho^V_k-\rho^V_k}} \\
  &= \sum_{j,j'} \p{q^V_{kj}}^* q^V_{kj'}\,
  \bbk{\epsilon_j \epsilon_{j'}} \\
  &\le \sum_j \abs{q^V_{kj}}^2 \epsilon^2
  = \p{\f{\epsilon}{M^V_k}}^2.
\end{align}
Using the fact that the operators $Q^V_k$ are orthonormal, we can therefore bound the mean squared 2-norm (Euclidean) distance between $\tilde\rho_V$ and $\rho$ as
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}_2^2}
  = \sum_k \Bbk{\abs{\tilde\rho^V_k - \rho^V_k}^2}
  \le \epsilon^2 \S_V^2,
  \label{eq:bound}
\end{align}
where the {\it error scale} $\S_V$ of $V$, defined by
\begin{align}
  \S_V \equiv \norm{M_V^{-1}}_2 = \sqrt{\sum_k\f1{\p{M^V_k}^2}},
  \label{eq:scale}
\end{align}
is the 2-norm of the pseudoinverse $M_V^{-1}$ of $M_V$.
Note that the bound in Eq.~\eqref{eq:bound} only holds if the measurement matrix $M_V$ is nonsingular, as otherwise some of the operators $Q^V_k$ that we used to expand $\rho$ and $\tilde\rho_V$ are undefined, reflecting the fact that if $M_V$ is singular then measuring spin projections along all axes in $V$ does not provide sufficient information to reconstruct arbitrary quantum states.

The complexity of computing the error scale $\S_V$ is determined by the need to construct the measurement matrix $M_V$, and compute its singular values.
These computations can be drastically reduced by use of the transition operators.
Specifically, the orthogonality of transition operators with different degrees, i.e.~$\obk{T_{\v v\ell,0}|T_{\v v',\ell',0}} = 0$ for $\ell\ne\ell'$, implies that $M_V$ has a block-diagonal structure in the transition operator basis, with rectangular blocks of size $\p{2\ell+1}\times\abs{V}$:
\begin{align}
  M_{V\ell} \equiv U_\ell \sum_{\v v} \oket{T_{\v v\ell,0}} \bra{\v v},
\end{align}
where $U_\ell$ is a $\p{2\ell+1}\times d^2$ isometry defined by
\begin{align}
  U_\ell \equiv \sum_{m=-\ell}^\ell \ket{m} \obra{T_{\ell m}},
\end{align}
with $\obra{T_{\ell m}}$ the conjugate transpose of $\oket{T_{\ell m}}$.
Using the blocks $M_{V\ell}$, we can expand
\begin{align}
  \S_V^2 = \sum_\ell \S_{V\ell}^2,
  &&
  \S_{V\ell} \equiv \norm{M_{V\ell}^{-1}}_2,
\end{align}
where $M_{V\ell}^{-1}$ is the pseudoinverse of $M_{V\ell}$.
Altogether, the expansion $T_{\v v\ell,0} = R\p{\v v} T_{\ell,0} R\p{\v v}^\dag$ enables the construction of $M_{V\ell}$ in $O(\abs{V}d^3)$ time, and computing the singular values of $M_{V\ell}$ takes at most $O(\abs{V}d^2)$ time \footnote{The calculation of $\S_{V\ell}$ can be sped up by the fact that its value is independent of $d$ (as long as $d\ge\ell+1$), so we can essentially set $d=\ell+1$ for the purposes of computing $\S_{V\ell}$}.
If we assume that $\abs{V}\sim d$, then computing $\S_V$ to certify the bound in Eq.~\eqref{eq:bound} takes $O(d^5)$ serial or $O(d^4)$ parallel runtime (see Figure [\red{REF}]).
[\red{todo: add figure showing runtimes as a function of $d$ with $\abs{V}=2d-1$; modify text appropriately after making the figure}].

[\red{todo: add a derivation of the error bound that assumes $n$ measurements of spin projection along each axis to Appendix \ref{sec:error}.
  Briefly discuss these results in this section.}]

@@

The reconstruction error bound in Eq.~\eqref{eq:bound} provides a straightforward answer: the number of measurement axes should be chosen to minimize the product $\epsilon^2\S_V^2$, which bounds the expected error in the estimate $\tilde\rho_V$ of an unknown state $\rho$.
If there are a total of $N$ spin projection measurements onto $r$ quantization axes, then $n=N/r$ measurements are used to estimate spin projection onto any given axis.
The variance in estimates of spin projection is then $\epsilon^2\sim1/n\sim r$ at a fixed total number of measurements $N$.
Altogether, the number of measurement axes should be chosen in such a way as to minimize $r\S_V^2$ with $\abs{V}=r$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A closer look at measurement error}
\label{sec:error}

[\red{todo: derive an error bound that relies on the number of spin projection measurements}]
